Universal acceptance of (nearly all of) basic science in the US

### 2024-06-24 paper.tex
making summary plot and adding some first introduction bits

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Who mistrust basic science?},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Who mistrust basic science?}
\author{\phantom{0}}
\date{}

\shorttitle{Who mistrust basic science?}

@@

There is increasingly more talk of defiance towards science and towards experts more generally. In every population, many people only trust science `some of the time', and a sizable minority doesn't trust it.

This appears difficult to reconcile with the fact that, in many countries, most people receive at least a basic science education, that science education (along with education more generally) is increasing, and that science education seems to be the main driver of trust in science.

But how deep is distrust of science? Do people reject science wholesale, or is it more driven by attitudes towards specific contents, maybe in particular those seen as being politicized, or as going against deeply held beliefs, or behaviors they want to engage in? {[}previous work suggests that issue-specific claims provoke more negative attitudes{]}

Here, we look at whether people in the US, across different levels of trust in science (including a sample with anti-vaccine attitudes), accept basic scientific facts.

\subsection{Levels of trust in science}\label{levels-of-trust-in-science}

\subsection{Why do people distrust science?}\label{why-do-people-distrust-science}

{[}here? We know from the many studies on the gateway belief model that people don't always accept the scientific consensus (indeed, the effect sizes are usually not massive){]}

\subsection{How much science do people know?}\label{how-much-science-do-people-know}

This is a crucial point: if there are questions with specific scientific answers that just about everyone already knows, then in a sense our point is already proven\ldots{} so the first thing to do would be to look at data on knowledge of science, and see what the highest rates of correct answer is, to see how much room for improvement there is.

@@

To measure conspiracy thinking, we selected 10 science/health related conspiracy theories from the Belief in Conspiracy Theory Inventory (BCTI) by Pennycook, Binnendyk, and Rand (2022) (Table \ref{tab:conspiracy}).

@@

\paragraph{Trust in science}\label{trust-in-science}

@@

\begin{figure}
\centering
\includegraphics{output/figures/exp1-plot.pdf}
\caption{\label{fig:exp1-plot}\textbf{A} Shows the distribution of science knowledge (left) and acceptance of scientific consensus \textbf{B} Shows the relationship between trust in science and science knowledge/acceptance of scientific consensus \textbf{C} Shows the relationship between conspiracy thinking and science knowledge/acceptance of scientific consensus}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{output/figures/exp2-plot.pdf}
\caption{\label{fig:exp2-plot}\textbf{A} Shows the distribution of science knowledge (left) and acceptance of scientific consensus for participants who gave the wrong answer \textbf{B} Shows the relationship between trust in science and science knowledge/acceptance (if wrong at first, rounded to the first digit) of scientific consensus \textbf{C} Shows the relationship between conspiracy thinking and science knowledge/acceptance (if wrong at first) of scientific consensus}
\end{figure}

@@

@@

\begin{figure}
\centering
\includegraphics{output/figures/exp3-plot.pdf}
\caption{\label{fig:exp3-plot}\textbf{A} Shows the distribution of science knowledge (left) and acceptance of scientific consensus for participants who gave the wrong answer \textbf{B} Shows the relationship between trust in science and science knowledge/acceptance (if wrong at first, rounded to the first digit) of scientific consensus \textbf{C} Shows the relationship between conspiracy thinking and science knowledge/acceptance (if wrong at first) of scientific consensus}
\end{figure}

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-bruderMeasuringIndividualDifferences2013}
Bruder, M., Haffke, P., Neave, N., Nouripanah, N., \& Imhoff, R. (2013). Measuring individual differences in generic beliefs in conspiracy theories across cultures: Conspiracy mentality questionnaire. \emph{Frontiers in Psychology}, \emph{4}. Retrieved from \url{https://www.frontiersin.org/articles/10.3389/fpsyg.2013.00225}

@@

@@

@@

@@

\subsubsection{Trust in science}\label{trust-in-science-1}

@@

@@

\subsection{Trust in science}\label{trust-in-science-3}

@@

\begin{longtable}[t]{>{}r>{\raggedright\arraybackslash}p{40em}}
\caption{\label{tab:unnamed-chunk-29}Reasons for consensus rejection}\\
\toprule
id & comments\\
\midrule
1 & no\\
20 & None\\
22 & I answered thoughtfully, good luck!\\
27 & interesting, thank you.\\
30 & n/a\\
\addlinespace
32 & None\\
39 & no\\
41 & none\\
42 & n/a\\
67 & thanks\\
\addlinespace
69 & No comments\\
74 & no\\
76 & none\\
78 & None - thank you!\\
79 & Should've clicked agree about plants getting carbon from the air, but I wanted to look it up first. Whenever I encounter something I didn't know online, I like to cross-reference. In retrospect though, everything else you asked about was simple and on the level, and I'm sure you are correct.

@@

@@

See Appendix \ref{exp1}.

@@

\subsubsection{Trust in science}\label{trust-in-science-4}

@@

\subsection{Trust in science}\label{trust-in-science-5}

@@

@@

\subsubsection{Trust in science}\label{trust-in-science-6}

@@

\subsection{Trust in science}\label{trust-in-science-7}

@@

>>>>

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Universal acceptance of (nearly all of) basic science in the US},
  pdflang={en-EN},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Universal acceptance of (nearly all of) basic science in the US}
\author{\textsuperscript{}}
\date{}

\shorttitle{Acceptance of basic science}

@@

How much do people trust and accept science? The recent COVID pandemic has highlighted the importance of this question, with trust in science being the best predictor of people's proclivity to follow health guidelines (Algan, Cohen, Davoine, Foucault, \& Stantcheva, 2021).

Unfortunately, trust in science is far from being at ceiling (Wellcome Global Monitor, 2018, 2020), and there are suggestions that it has recently been dropping in some countries (Algan et al., 2021) and that it is increasingly politicized in the US (Gauchat, 2012; Krause, Brossard, Scheufele, Xenos, \& Franke, 2019; Li \& Qian, 2022).

This relative lack of trust in science could mean different things. It could mean that some people reject science wholesale, or at least to a significant extent. Or it could mean that they maintain a high degree of trust in most of science, and only question specific results that happen to contradict specific other beliefs they have (creationism, etc.), or beliefs that are associated with disliked behaviors (vaccination, acting for climate change).

We answer this question by looking at how much Americans accept basic scientific knowledge. We'll pay special attention to two groups of participants: people who say they don't trust science (much, or at all), and people who endorse science-related conspiracy theories (CTs), or have a CT mindset, since this has been linked with a low degree of trust in science, and that many CTs imply a mistrust of at least some science (vaccination, hydroxychloroquine, etc.).

We presently briefly review work on trust in science, and science knowledge, before turning to work on CT and trust in science.

\subsection{Trust in science}\label{trust-in-science}

Across the globe, most people say they trust science at least to some extent. In 2018, the Wellcome Global Monitor (WGM) surveyed of over 140'000 people in over 140 countries on trust in science (Wellcome Global Monitor, 2018) . In 2020, during the first year of the Covid pandemic, a follow up survey was made in 113 countries, involving 119'000 participants (Wellcome Global Monitor, 2020). Across all countries, in 2018, 32\% of participants said they trust science ``a lot'' (41\% in 2020), 45\% trust ``some'' science (39\% in 2020), and only 13\% percent trust science ``not much/ not at all'' (13\% in 2020), while 10\% indicated ``don't know'' (7\% in 2020) (\url{https://wellcome.org/reports/wellcome-global-monitor-covid-19/2020\#gid=6acd&pid=0}).

Trust in science also appears to be relatively stable in the US (Funk \& Kennedy, 2020). However, there are some potentially worrying trends: The US might be increasingly polarized on trust in science (Gauchat, 2012). For example, the partisan divide on the question of whether climate change is a major threat to societey \href{https://www.pewresearch.org/short-reads/2023/08/09/what-the-data-says-about-americans-views-of-climate-change/}{has been growing in the last decade} (Kennedy, 2023). And while by some measures only about half of the US population believes in anthropogenic climate change, people do mostly not seem to change their mind on on the issue (Motta, 2021).

Trust in science is typically measured in three ways: either asking explictly about general trust in science or scientists, by asking about science attitudes (e.g.~XX), or asking questions about specific, typically contentious, science topics (e.g.~vaccines, climate change, GMOs). It is unclear to which extent these measures capture trust in basic science. To assess this, we can look at the literature on science knowledge.

\subsection{Science knowledge}\label{science-knowledge}

A lot of research has been conducted on the public understanding of science. Part of this research are surveys measuring to which extent people have basic science knowledge (e.g.~``Does the Earth go around the Sun, or does the Sun go around the Earth?''). For us, these surveys provide a lower bound to how much people trust basic science: people who give the right answer (and not by chance) supposedly trust science on that question. Unfortunately, that lower bound has often been indeed pretty low: researchers have pointed out the deficit in basic science knowledge (Miller, 1998, 2004). For example, only 55\% in the US (in 2014) and only 46\% in the EU (in 2005) knew that antibiotics kill only bacteria but not viruses (Science Literacy et al., 2016). A question that remains is whether people do not provide the right answers because they are not familiar with the scientific consensus, or because they explicitly reject that consensus.

{[}To do: Make a descriptive overview of science knowledge across time/countries. This is a crucial point: if there are questions with specific scientific answers that just about everyone already knows, then in a sense our point is already proven. But there are enough questions where people's knowledge appears quite poor anyways{]}

\subsection{CT and TiS}\label{ct-and-tis}

Recently, the relationship of one group to trust in science (TiS) has been particularly scrutinized: conspiracy theorists (CTists), either defined as people who endorse specific conspiracy theories, or as people who have a conspiracy mindset (Rutjens \& Većkalov, 2022; Vranic, Hromatko, \& Tonković, 2022). CTists reject specific science knowledge (related to their CT), and they tend to trust science less on the whole. What remains unclear is to which extent CTists (dis)trust basic science.

\subsection{The present studies}\label{the-present-studies}

In a series of studies, we ask basic science knowledge questions, inform people of the scientific consensus, and see if they accept it. Of particular interest are people who are low in TiS, and CTists.

In study 1, we plainly inform participants of the scientific consensus and ask if they accept it. We use a subset of questions on basic science knowledge that have been used in numerous public opinion surveys (Allum, Sturgis, Tabourazi, \& Brunton-Smith, 2008; Durant, Evans, \& Thomas, 1989; Miller, 1998) sometimes referred to as the ``Oxford scale'' (Gauchat, 2011). In study 2, we remove a problematic science question and present people with a more elaborated explanation of the scientific consensus, as well as additional sources. This allows us to remove the issue that participants might simply not have trusted us to correctly report the scientific consensus in study 1. Study 3 follows essentially the same design as study 2, but is run on a vaccine-skeptic sample. Further, we provided participants with an explicit option to revise their answer, in case they rejected the consensus. We also asked follow-up questions on why participants accepted the scientific consensus, namely whether they thought it was because they trust scientists or because they verified independently. Study 4 uses the same design as study 3, but asking science questions that people are less likely to have encountered before, and that they would be less likely to be able to verify (or even understand) themselves {[}on an anti-vaxx sample again?{]}.

Then: replications in other countries?

@@

To measure conspiracy thinking, we selected 10 science/health related conspiracy theories from the Belief in Conspiracy Theory Inventory (BCTI) by Pennycook, Binnendyk, and Rand (2022) (Table \ref{tab:conspiracy}). Participants were asked: ``Below is a list of events for which the official version has been disputed. For each event, we would like you to indicate to what extent you believe the cover-up version of events is true or false. {[}1-9; labels: 1 - completly false, 5 - unsure, 9 - completely true{]}''.

@@

\paragraph{Trust in science}\label{trust-in-science-1}

@@

@@

@@

\begin{figure}
\centering
\includegraphics{output/figures/summary-plot.pdf}
\caption{\label{fig:summary-plot}Points represent the average share of acceptance as a function of the \textbf{A} level of trust in science (``In general, would you say that you trust science a lot, some, not much, or not at all? {[}1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot{]}''); \textbf{B} average conspiracy thinking (CMQ, 5 items on a scale from 0 to 100); \textbf{C} belief in a specific conspiracy theory. Participants were asked to rate their belief in the conspiracy on a scale from 1 to 9, with the labels: ``1 - completly false, 5 - unsure, 9 - completely true''. Averages are only based on participants who believed at least to some degree in the conspiracy theory. We consider a believer as anyone scoring higher than the scale midpoint, i.e.~\textgreater5. Number labels in plots represent the number of participants for the corresponding data point.}
\end{figure}

@@

\section{Discussion}\label{discussion-2}

There is increasingly more talk of defiance towards science and towards experts more generally. In every population, many people only trust science `some of the time', and a sizable minority doesn't trust it. This appears difficult to reconcile with the fact that, in many countries, most people receive at least a basic science education, that science education (along with education more generally) is increasing, and that science education seems to be the main driver of trust in science.

But how deep is distrust of science? In four studies, we have shown that almost every participant accepts almost all of basic science questions we included.

Traditionally, people have interpreted people's lack of basic knowledge in light of the deficit-model. Here, we show that people readily accept the scientifically consensual answer to these questions.

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-alganTrustScientistsTimes2021}
Algan, Y., Cohen, D., Davoine, E., Foucault, M., \& Stantcheva, S. (2021). Trust in scientists in times of pandemic: Panel evidence from 12 countries. \emph{Proceedings of the National Academy of Sciences}, \emph{118}(40), e2108576118. \url{https://doi.org/10.1073/pnas.2108576118}

\bibitem[\citeproctext]{ref-allumScienceKnowledgeAttitudes2008}
Allum, N., Sturgis, P., Tabourazi, D., \& Brunton-Smith, I. (2008). Science knowledge and attitudes across cultures: a meta-analysis. \emph{Public Understanding of Science}, \emph{17}(1), 35--54. \url{https://doi.org/10.1177/0963662506070159}

\bibitem[\citeproctext]{ref-bruderMeasuringIndividualDifferences2013}
Bruder, M., Haffke, P., Neave, N., Nouripanah, N., \& Imhoff, R. (2013). Measuring individual differences in generic beliefs in conspiracy theories across cultures: Conspiracy mentality questionnaire. \emph{Frontiers in Psychology}, \emph{4}. Retrieved from \url{https://www.frontiersin.org/articles/10.3389/fpsyg.2013.00225}

@@

\bibitem[\citeproctext]{ref-durantPublicUnderstandingScience1989}
Durant, J. R., Evans, G. A., \& Thomas, G. P. (1989). The public understanding of science. \emph{Nature}, \emph{340}(6228), 11--14. \url{https://doi.org/10.1038/340011a0}

\bibitem[\citeproctext]{ref-funkPublicConfidenceScientists2020}
Funk, C., \& Kennedy, B. (2020). \emph{Public confidence in scientists has remained stable for decades}.

\bibitem[\citeproctext]{ref-gauchatCulturalAuthorityScience2011}
Gauchat, G. (2011). The cultural authority of science: Public trust and acceptance of organized science. \emph{Public Understanding of Science}, \emph{20}(6), 751--770. \url{https://doi.org/10.1177/0963662510365246}

\bibitem[\citeproctext]{ref-gauchatPoliticizationSciencePublic2012}
Gauchat, G. (2012). Politicization of Science in the Public Sphere: A Study of Public Trust in the United States, 1974 to 2010. \emph{American Sociological Review}, \emph{77}(2), 167--187. \url{https://doi.org/10.1177/0003122412438225}

\bibitem[\citeproctext]{ref-kennedyWhatDataSays2023}
Kennedy, A. T. C. F. and B. (2023). \emph{What the data says about americans{'} views of climate change}. Retrieved from \url{https://www.pewresearch.org/short-reads/2023/08/09/what-the-data-says-about-americans-views-of-climate-change/}

\bibitem[\citeproctext]{ref-krauseTrendsAmericansTrust2019a}
Krause, N. M., Brossard, D., Scheufele, D. A., Xenos, M. A., \& Franke, K. (2019). Trends{\textemdash}americans{'} trust in science and scientists. \emph{Public Opinion Quarterly}, \emph{83}(4), 817--836. \url{https://doi.org/10.1093/poq/nfz041}

@@

\bibitem[\citeproctext]{ref-liPolarizationPublicTrust2022}
Li, N., \& Qian, Y. (2022). Polarization of public trust in scientists between 1978 and 2018: Insights from a cross-decade comparison using interpretable machine learning. \emph{Politics and the Life Sciences}, \emph{41}(1), 45--54. \url{https://doi.org/10.1017/pls.2021.18}

\bibitem[\citeproctext]{ref-millerMeasurementCivicScientific1998a}
Miller, J. D. (1998). The measurement of civic scientific literacy. \emph{Public Understanding of Science}, \emph{7}(3), 203--223. \url{https://doi.org/10.1088/0963-6625/7/3/001}

\bibitem[\citeproctext]{ref-millerPublicUnderstandingAttitudes2004}
Miller, J. D. (2004). Public Understanding of, and Attitudes toward, Scientific Research: What We Know and What We Need to Know. \emph{Public Understanding of Science}, \emph{13}(3), 273--294. \url{https://doi.org/10.1177/0963662504044908}

\bibitem[\citeproctext]{ref-mottaChangingMindsChanging2021}
Motta, M. (2021). Changing Minds or Changing Samples? Disentangling Microlevel Stability and Macrolevel Growth in Anthropogenic Climate Change Beliefs. \emph{International Journal of Public Opinion Research}, \emph{33}(2), 477--489. \url{https://doi.org/10.1093/ijpor/edaa020}

@@

\bibitem[\citeproctext]{ref-rutjensConspiracyBeliefsScience2022}
Rutjens, B. T., \& Većkalov, B. (2022). Conspiracy beliefs and science rejection. \emph{Current Opinion in Psychology}, \emph{46}, 101392. \url{https://doi.org/10.1016/j.copsyc.2022.101392}

\bibitem[\citeproctext]{ref-committeeonscienceliteracyandpublicperceptionofscienceScienceLiteracyConcepts2016}
Science Literacy, C. on, Public Perception of Science, Board on Science Education, Behavioral, D. of, Sciences, S., Education, \& National Academies of Sciences, Engineering, and Medicine. (2016). \emph{Science Literacy: Concepts, Contexts, and Consequences} (C. E. Snow \& K. A. Dibner, Eds.). Washington, D.C.: National Academies Press. \url{https://doi.org/10.17226/23595}

\bibitem[\citeproctext]{ref-vranicDidMyOwn2022}
Vranic, A., Hromatko, I., \& Tonković, M. (2022). {``}I did my own research{''}: Overconfidence, (dis)trust in science, and endorsement of conspiracy theories. \emph{Frontiers in Psychology}, \emph{13}. \url{https://doi.org/10.3389/fpsyg.2022.931865}

\bibitem[\citeproctext]{ref-wellcomeglobalmonitorWellcomeGlobalMonitor2018}
Wellcome Global Monitor. (2018). \emph{Wellcome Global Monitor 2018}. Retrieved from \url{https://wellcome.org/reports/wellcome-global-monitor/2018}

\bibitem[\citeproctext]{ref-wellcomeglobalmonitorWellcomeGlobalMonitor2020}
Wellcome Global Monitor. (2020). \emph{Wellcome Global Monitor 2020: Covid-19}. Retrieved from \url{https://wellcome.org/reports/wellcome-global-monitor-covid-19/2020}

@@

@@

\subsubsection{Trust in science}\label{trust-in-science-3}

@@

\subsection{Trust in science}\label{trust-in-science-4}

@@

\begin{longtable}[t]{>{}r>{\raggedright\arraybackslash}p{40em}}
\caption{\label{tab:unnamed-chunk-21}Reasons for consensus rejection}\\
\toprule
id & comments\\
\midrule
1 & no\\
20 & None\\
22 & I answered thoughtfully, good luck!\\
27 & interesting, thank you.\\
30 & n/a\\
\addlinespace
32 & None\\
39 & no\\
41 & none\\
42 & n/a\\
67 & thanks\\
\addlinespace
69 & No comments\\
74 & no\\
76 & none\\
78 & None - thank you!\\
79 & Should've clicked agree about plants getting carbon from the air, but I wanted to look it up first. Whenever I encounter something I didn't know online, I like to cross-reference. In retrospect though, everything else you asked about was simple and on the level, and I'm sure you are correct.

@@

\subsection{Additional plots}\label{additional-plots}

\begin{figure}
\centering
\includegraphics{output/figures/exp1-plot-overview.pdf}
\caption{\label{fig:exp1-plot-overview}\textbf{A} Shows the distribution of science knowledge (left) and acceptance of scientific consensus \textbf{B} Shows the relationship between trust in science and science knowledge/acceptance of scientific consensus \textbf{C} Shows the relationship between conspiracy thinking and science knowledge/acceptance of scientific consensus}
\end{figure}

@@

@@

\subsubsection{Trust in science}\label{trust-in-science-5}

@@

\subsection{Trust in science}\label{trust-in-science-6}

@@

\subsection{Additional plots}\label{additional-plots-1}

@@

\subsubsection{Trust in science}\label{trust-in-science-7}

@@

\subsection{Trust in science}\label{trust-in-science-8}

@@

\subsection{Additional plots}\label{additional-plots-2}

\begin{figure}
\centering
\includegraphics{output/figures/exp3-plot-overview.pdf}
\caption{\label{fig:exp3-plot-overview}\textbf{A} Shows the distribution of science knowledge (left) and acceptance of scientific consensus for participants who gave the wrong answer \textbf{B} Shows the relationship between trust in science and science knowledge/acceptance (if wrong at first, rounded to the first digit) of scientific consensus \textbf{C} Shows the relationship between conspiracy thinking and science knowledge/acceptance (if wrong at first) of scientific consensus}
\end{figure}
